---
title: "Introduction to R package fastfrechet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R package fastfrechet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fastfrechet)
```

# Introduction

`fastfrechet` is an R package providing fast and robust solutions to several
computational problems associated with Fréchet regression (Petersen et al. 2019)
and an associated variable selection method (Tucker et al. 2023), focusing on a
specific metric space setting. Fréchet regression generalizes regression with
Euclidean covariates and Euclidean responses, to settings where responses exist
in a general metric space (i.e. non-vector space). The space of univariate
distribution response objects is one such example, recently gaining popularity
in biomedical applications where technological advances have made rich
patient-level data sets commonplace in a variety of settings. These larger data
sets and complex response constraints require newer, faster algorithms for
feasibility and scalability (Coulter et al. 2024), which `fastfrechet` provides.

This vignette walks through an example use case of all of the available
functions in `fastfrechet`. This includes simulating example covariate-dependent
distribution responses, implementing Fréchet regression, performing variable
selection, and utilizing resampling tools to supplement automatic variable
selection. Where appropriate, the mathematical problems being solved by each
function are briefly described. A second vignette, `monotoneQP-fastfrechet`,
provides a more detailed algorithmic exposition of the function `monotoneQP`
specifically.


# Running Through an Example

## Generating quantile functions

### Using `fastfrechet::generate_zinbinom_qf`

### Plotting


## Fréchet regression

### Using `fastfrechet::frechetreg_univar2wass`

### Plotting $\widehat{\mathbf{Y}}$ and $\widehat{\mathbf{Q}}$


## Variable selection with Fréchet ridge selection operator

### Using `fastfrechet::FRiSO_univar2wass`

### Plotting $\widehat{\pmb{\lambda}}(\tau)$ solution paths



# Resampling for Variable Selection

## Cross-validation

### Using `fastfrechet::FRiSO_CV_univar2wass`

### Plotting, interpreting CV errors and optimal solution

## Complementary pairs stability selection

### Using `fastfrechet::FRiSO_CPSS_univar2wass`

### Plotting, interpreting stability paths

### Shah and Samworth error control