---
title: "Introduction to R package fastfrechet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to R package fastfrechet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(fastfrechet)
```

# Introduction



# Running Through an Example

## Generating quantile functions

### Using `fastfrechet::generate_zinbinom_qf`

### Plotting


## Fréchet regression

### Using `fastfrechet::frechetreg_univar2wass`

### Plotting $\widehat{\mathbf{Y}}$ and $\widehat{\mathbf{Q}}$


## Variable selection with Fréchet ridge selection operator

### Using `fastfrechet::FRiSO_univar2wass`

### Plotting $\widehat{\pmb{\lambda}}(\tau)$ solution paths



# Resampling for Variable Selection

## Cross-validation

### Using `fastfrechet::FRiSO_CV_univar2wass`

### Plotting, interpreting CV errors and optimal solution

## Complementary pairs stability selection

### Using `fastfrechet::FRiSO_CPSS_univar2wass`

### Plotting, interpreting stability paths

### Shah and Samworth error control