---
title: "Overview of Dual Active-Set Method for monotoneQP"
output: rmarkdown::html_vignette
bibliography: '`r system.file("REFERENCES.bib", package="fastfrechet")`'
vignette: >
  %\VignetteIndexEntry{Overview of Dual Active-Set Method for monotoneQP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

This vignette details the mathematical foundation of the algorithm we implement to solve the non-convex Fréchet mean problem for univariate distribution responses equipped with the 2-Wasserstein metric. In the "population setting", this is an orthogonal projection problem. In the "sample setting" where distribution responses are represented by quantile functions $\mathbf{y} \in \mathbb{R}^m$ evaluated on a discrete uniform $m$-grid in $(0, 1)$ and obeying box constraints $y_j \in [b_L, b_U] \subseteq \mathbb{R}$, the Fréchet mean problem can be written

$$
\widehat{\mathbf{q}} := \mathrm{arg\:min}_{\mathbf{q} \in \mathbb{R}^m} \lVert{\mathbf{q} - \widehat{\mathbf{y}}} \rVert_2^2, \qquad \mathbf{b} - \mathbf{A}^{\top}\mathbf{q} \leq \pmb{0}_{m+1},
$$
where $\mathbf{b}^{\top} = \begin{pmatrix} b_L & \pmb{0}_{m-1}^{\top} & -b_U\end{pmatrix} \in \mathbb{R}^{m+1}$,
$\mathbf{A} \in \mathbb{R}^{m\times(m+1)}$ is a first-difference matrix of the form

$$
\mathbf{A} = \begin{pmatrix}
  +1 & -1 &  0 &  0 & \cdots &  0 &  0 &  0 \\
   0 & +1 & -1 &  0 & \cdots &  0 &  0 &  0 \\
   0 &  0 & +1 & -1 & \cdots &  0 &  0 &  0 \\
   0 &  0 &  0 & +1 & \cdots &  0 &  0 &  0 \\
   \vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots & \vdots \\
   0 &  0 &  0 &  0 & \cdots & +1 & -1 &  0 \\
   0 &  0 &  0 &  0 & \cdots &  0 & +1 & -1
\end{pmatrix},
$$
and finally $\widehat{\mathbf{y}}$ is a non-convex sum of the $\mathbf{y}$'s
which does not necessarily sit in the target space. The final projected vector $\widehat{\mathbf{q}}$ is monotone non-decreasing (i.e. $q_i \leq q_j$ for $i \leq j$) and obeys the same box constraints as the $\mathbf{y}$'s. The `fastfrechet` package implements a solver for this problem via the `monotoneQP` function.

## Example Utilization

To first visualize what `monotoneQP` does, we specify a set of box constraints $b_L = 0.5$ and $b_U = 1.5$, and generate an example vector which does not obey those constraints and does not exhibit monotonicity.
```{r}
# Set box constraints:
lower = 0.5
upper = 1.5

# Generate example vector:
m = 100

set.seed(31)
y = rnorm(m, seq(0, 2, len = m), 0.1)
```

We next calculate the fitted vector...
```{r}
# Calculate monotone, box-constrained projection:
output = fastfrechet::monotoneQP(y, lower = 0.5, upper = 1.5)
```
which fits inside the box constraints...
```{r}
# ...bounded from below by the lower bound...
min(output$Solution) - lower

# ...and bounded from above by the upper bound...
max(output$Solution) - upper
```
and has monotone non-decreasing entries.
```{r}
min(diff(output$Solution[1, ]))
```
The original vector and the fitted vector are plotted together.
```{r, fig.width = 6.25, fig.height = 5}
# Plot values of the generated vector:
plot(y, main = "Monotone and Box Constrained", las = 1,
     xlab = "Vector Entry", ylab = expression("y, "*hat("q")))
abline(h = c(lower, upper), lty = 2, col = "gray80")

# Add values of monotone projection:
points(output$Solution[1, ], pch = 20)

legend("topleft", pch = c(1, 20), bty = "n",
       legend = c("Unconstrained", "Monotone Constrained"))
```

## Algorithm

Solving for $\widehat{\mathbf{q}}$ is a quadratic programming (QP) problem, for which several approaches and solvers exist. We introduce a modification of the dual active-set method of Arnström et al. (2022, Algorithm 1) which utilizes the sparse structure of $\mathbf{A}$ to avoid costly matrix multiplication and $LDL^{\top}$ decomposition steps. This algorithm is implemented in *C++* as a sub-routine in `fastfrechet` functions solving Fréchet mean problems for 2-Wasserstein space, and is also made available as a standalone QP solver in the `monotoneQP` function.

### Notation

The dual active-set algorithm of @arnstrom_dual_2022 is for QP problems with general constraints. We translate it into our context with the following notation. Denote $[m] := \{1, \dots, m\}$, and let $\mathbf{d} := \mathbf{A}^{\top}\mathbf{y} - \mathbf{b}$. For iteration $k \geq 0$, let $W_k \subseteq \{1, \dots, m + 1\}$ be an index set in increasing order, and let $\bar{W}_k = [m+1] \setminus W_k$ be its similarly-ordered relative complement in $[m+1]$. Denote with $\mathbf{A}_k$ the sub-matrix of $\mathbf{A}$ consisting of the columns indexed by $W_k$ (when defined), and denote with $\bar{\mathbf{A}}_k$ the sub-matrix consisting of the remaining columns (also when defined). Similarly, let $\mathbf{d}_k$ be a vector consisting of the elements of $\mathbf{d}$ indexed by $W_k$, and $\bar{\mathbf{d}}_k$ be a vector consisting of the remaining elements.

### The algorithm

The following algorithm variables are updated each iteration.

* $\pmb{\eta}_k \rightarrow \pmb{\eta}^* \in \mathbb{R}^{m+1}$, the Lagrange
multiplier associated with the primal monotone QP problem.
* $\pmb{\mu}_k \rightarrow \pmb{\mu}^* \in \mathbb{R}^{m+1}$, the Lagrange
multiplier associated with the dual problem to the monotone QP problem.
* $W_k$ ($\bar{W}_k$), the working index set identifying which entries of $\pmb{\eta}_k$ may be positive (must be zero).

The full Algorithm 1 from @arnstrom_dual_2022 is given below, with applicable variable substitutions for our context. We also take advantage of two important facts about $\mathbf{A}$: first, $\mathbf{A}$ is full row-rank, but not full column rank; second, any sub-matrix of $\mathbf{A}$ defined by taking a strict subset of its columns is full column-rank.

1. Initialize $k=0$, $\mathbf{d} = \mathbf{A}^{\top}\mathbf{y} - \mathbf{b}$,
$\pmb{\eta}_0 = \pmb{0}_{m+1}$, $W_0 = \emptyset$, $\pmb{\mu}_0 =\pmb{0}_{m+1}$,
and $\mathbf{exit} = \verb+false+$.
2. **while** !$\mathbf{exit}$ **do**
    a. **if** $W_k \neq [m + 1]$ **then**
        * $[\tilde{\pmb{\eta}}_k]_{W_k} \leftarrow -(\mathbf{A}_k^{\top}\mathbf{A}_k)^{-1}\mathbf{d}_k$.
        * **if** $\tilde{\pmb{\eta}}_k \geq \pmb{0}$ **then**
            - $[\pmb{\mu}_k]_{\bar{W}_k} \leftarrow \bar{\mathbf{A}}{}^{\top}_k\mathbf{A}_k [\tilde{\pmb{\eta}}_k]_{W_k} + \bar{\mathbf{d}}_k$.
            - $\pmb{\eta}_{k+1} \leftarrow \tilde{\pmb{\eta}}_k$.
            - **if** $\pmb{\mu}_k \geq \pmb{0}$ **then**
                + $\mathbf{exit} \leftarrow \verb+true+$.
            - **else**
                + $j \leftarrow \mathrm{arg\:min}_{i \in \bar{W}_k} [\pmb{\mu}_k]_i$.
                + $W_{k+1} \leftarrow \mathrm{sort}(W_k \cup \{j\})$.
                + $\bar{W}_{k+1} \leftarrow \bar{W}_k \setminus \{j\}$.
        * **else**
            - $\mathbf{p}_k \leftarrow \tilde{\pmb{\eta}}_k - \pmb{\eta}_k$.
            - $B \leftarrow \{i \in W_k : [\tilde{\pmb{\eta}}_k]_i < 0\}$.
            - $j \leftarrow \mathrm{arg\:min}_{i \in B} -\frac{[\pmb{\eta}_k]_i}{[\mathbf{p}]_i}$.
            - $W_{k+1} \leftarrow W_k \setminus \{j\}$.
            - $\bar{W}_{k+1} \leftarrow \mathrm{sort}(\bar{W}_k \cup \{j\})$.
            - $\pmb{\eta}_{k+1} \leftarrow \pmb{\eta}_k - \frac{[\pmb{\eta}_k]_j}{[\mathbf{p}_k]_j} \cdot \mathbf{p}_k$.
    b. **else**
        * $\mathbf{p}_k \leftarrow$ solution to $\mathbf{A}^{\top}\mathbf{A}\mathbf{p}_k = \pmb{0}$, subject to $\mathbf{p}_k^{\top}\mathbf{d} < 0$.
        * $B \leftarrow \{i : [\mathbf{p}_k]_i < 0\}$.
        * $j \leftarrow \mathrm{arg\:min}_{i \in B} -\frac{[\pmb{\eta}_k]_i}{[\mathbf{p}_k]_i}$.
        * $W_{k+1} \leftarrow W_k \setminus \{j\}$.
        * $\bar{W}_{k+1} \leftarrow \mathrm{sort}(\bar{W}_k \cup \{j\})$.
        * $\pmb{\eta}_{k+1} \leftarrow \pmb{\eta}_k - \frac{[\pmb{\eta}_k]_j}{[\mathbf{p}_k]_j} \cdot \mathbf{p}_k$.
    c. $k \leftarrow k + 1$.
3. $W \leftarrow W_k$, $\pmb{\eta}^* \leftarrow \pmb{\eta}_k$, and $\widehat{\mathbf{q}} \leftarrow \mathbf{y} + \mathbf{A}_W\pmb{\eta}^*_W$.

### Matrix Simplifications

The implementation of the dual active-set method from @arnstrom_dual_2022 
involves iterative updates of the $LDL^{\top}$ decomposition of the
$\mathbf{A}_k^{\top}\mathbf{A}_k$ matrix. The form of $\mathbf{A}$ allows us to avoid these decompositions entirely, and represent the matrix operations $-(\mathbf{A}_k^{\top}\mathbf{A}_k)^{-1}\mathbf{d}_k$ and $\bar{\mathbf{A}}{}^{\top}_k\mathbf{A}_k [\tilde{\pmb{\eta}}_k]_{W_k} + \bar{\mathbf{d}}_k$ with a simpler set of subtraction operations. Further, in steps subsequent to 2(b) above, the result is invariant to any choice of $\mathbf{p}_k$ satisfying the two given equations. The upshot is the matrix $\mathbf{A}$ does not need to be stored, nor do expressions involving it or its sub-matrices need to be evaluated explicitly. We give a brief overview of each simplification, but omit the complete details.

#### Evaluating $-(\mathbf{A}_k^{\top}\mathbf{A}_k)^{-1}\mathbf{d}_k$

The matrix $\mathbf{A}^{\top}\mathbf{A}$ is tridiagonal of the form
$$
\mathbf{A}^{\top}\mathbf{A} = \begin{pmatrix}
  +1 & -1 &  0 &  0 & \cdots &  0 &  0 &  0 \\
  -1 & +2 & -1 &  0 & \cdots &  0 &  0 &  0 \\
   0 & -1 & +2 & -1 & \cdots &  0 &  0 &  0 \\
   0 &  0 & -1 & +2 & \cdots &  0 &  0 &  0 \\
   \vdots & \vdots & \vdots & \vdots & \ddots & \vdots & \vdots & \vdots \\
   0 &  0 &  0 &  0 & \cdots & +2 & -1 &  0 \\
   0 &  0 &  0 &  0 & \cdots & -1 & +2 & -1 \\
   0 &  0 &  0 &  0 & \cdots &  0 & -1 & +1
\end{pmatrix}.
$$
For $W_k \subset [m + 1]$ (strict subset), the matrix
$\mathbf{A}_k^{\top}\mathbf{A}_k$ comprises the entries at the intersections of
the columns/rows indexed by $W_k$, and so is also tridiagonal with possibly
isolated diagonal entries. We implement a solver which incrementally searches
along the indices in $W_k$ to identify isolated diagonal entries in
$\mathbf{A}_k^{\top}\mathbf{A}_k$ (given by indices $w \in W_k$ where
$w+1 \notin W_k$ and $w-1\notin W_k$) where inversion reduces to a simple scalar
division operation, or fully block tridiagonal entries in
$\mathbf{A}_k^{\top}\mathbf{A}_k$ (given by sequences of indices
$(w, w+1, \dots, w + s) \subset W_k$) where we apply a simplified Thomas
algorithm that takes advantage of the repeated diagonal entries of
$\mathbf{A}^{\top}\mathbf{A}$.


#### Evaluating $\bar{\mathbf{A}}{}^{\top}_k\mathbf{A}_k [\tilde{\pmb{\eta}}_k]_{W_k} + \bar{\mathbf{d}}_k$

As a preliminary matter, if $W_k = \emptyset$, we evaluate this expression to simply $\mathbf{d}$. For $W_k \notin \{\emptyset, [m+1]\}$, the matrix multiplication in the expression can be reduced to two steps:

1. solving $\mathbf{u} \leftarrow \mathbf{A}_k [\tilde{\pmb{\eta}}_k]_{W_k}$, and
2. solving $\mathbf{v} \leftarrow \bar{\mathbf{A}}_k^{\top} \mathbf{u}$.

In each case, as $\mathbf{A}$ is a difference matrix, left-multiplication by
$\mathbf{A}_k$ or by $\bar{\mathbf{A}}_k^{\top}$ involves taking scalar
differences of appropriate entries in $\tilde{\pmb{\eta}}_k$ (dictated by
$W_k$), or in $\mathbf{u}$ (dictated by $\bar{W}_k$), respectively.

#### Evaluating $\mathbf{A}^{\top}\mathbf{A}\mathbf{p}_k = \pmb{0}$ subject to $\mathbf{p}_k^{\top}\mathbf{d} < 0$

Recalling $\mathbf{A}$ is full row rank, if $\mathbf{A}^{\top}\mathbf{A}\mathbf{p}_k = \pmb{0}$ then we can see
$$
\mathbf{A} \mathbf{p}_k = (\mathbf{A}\mathbf{A}^{\top})^{-1}\mathbf{A}\pmb{0} = \pmb{0},
$$
implying (by the form of $\mathbf{A}$) that
$[\mathbf{p}_k]_i = [\mathbf{p}_k]_{i+1}$ for all $i < m + 1$. But then
$\mathbf{p}_k = p \cdot \pmb{1}$ for some $p$, and the constraint is now written
$$
\mathbf{p}^{\top}\mathbf{d} < 0 \quad \iff \quad p \cdot \pmb{1}^{\top}\mathbf{d} < 0.
$$
Note that
$$
\begin{align}
\pmb{1}^{\top}\mathbf{d} &= \pmb{1}^{\top} \mathbf{A}^{\top} \mathbf{y} - \mathbf{1}^{\top}\mathbf{b} \\
&= \pmb{0}^{\top}\mathbf{y} - (b_L - b_U) \\
&= b_U - b_L,
\end{align}
$$
so as long as the box constraints are valid—i.e. $b_L < b_U$—we have
$\pmb{1}^{\top}\mathbf{d} > 0$ and thus $p < 0$. Fixing $p < 0$ then, we see
$$
\begin{align}
B &\leftarrow [m+1] \\
j &\leftarrow \mathrm{arg\:min}_i -\frac{[\pmb{\eta}_k]_i}{p} = \mathrm{arg\:min}_i [\pmb{\eta}_k]_i \\
W_{k+1} &\leftarrow [m+1] \setminus \{j\} \\
\bar{W}_{k+1} &\leftarrow \{j\} \\
\pmb{\eta}_{k+1} &\leftarrow \pmb{\eta}_k - [\pmb{\eta}_k]_j \cdot \pmb{1},
\end{align}
$$
which do not depend on choice of $p$.